---
title: "dplyr Example 1 - BLS Wide to Tidy"
author: "Jim Porzak"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dplyr Example 1 - BLS Wide to Tidy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The Bureau of Labor Statistics (the BLS) [data page](http://www.bls.gov/data/), has
many data sets and many ways to access them. For a research project we needed the
monthly employment figures by "metro areas." Using the _Multi-screen Data Search_ tool for _Employment, Hours, and Earnings - State and Metro Area_ we captured the non-farm employment from 1995 through 2015 as a tab seperated file. See the BLS _sm_ [file spec](http://download.bls.gov/pub/time.series/sm/sm.txt)

There is a bit clean-up to do to make it readable with `read_tsv()` from Hadley's readr package.

#### Read the file into the data frame `employment` and take a glimpse of the first and last  ten columns

```{r LoadRawBLS}
library(dplyr)
library(tidyr)
library(readr)
library(dplyrExamples)
library(stringr)
library(ggplot2)
library(lubridate)
fn <- system.file("extdata", "BLS_NonFarmEmploymentInAreas_1995_2015.tsv",
                  package = "dplyrExamples")
file_out <- "file_out.tsv"
BLS_CleanRawTextFile(fn, file_out)
employment <- read_tsv(file_out)
dim(employment)
glimpse(employment[1:10])
glimpse(employment[244:253])
```

There are two challenges with this data set:

1. The *Series_ID* needs to be decoded to pull out the state and area for the row. 
1. It is _very_ wide. We wish to tidy it up - in the Hadley sense.

The series_id decoder is in Section 5 of the [SM file spec](http://download.bls.gov/pub/time.series/sm/sm.txt) and is repeated here:

```
0        1         2    <--
12345678901234567890    <-- character counter 
SMU01266207072200001    <-- Sample Series ID

Positions Code		          			Value

1-2       survey abbreviation	=		SM
3         seasonal (code)   	=		U
4-5       state_code	      	=		01
6-10      area_code	        	=		26620
11-12     supersector_code  	=		70
13-18     industry_code	    	=		70722000
19-20     data_type_code  		=		01
```
We did not need NIAC's code breakdown so positions 11-18 are all zeros in our extract.

The BLS has standard lookup files for the state and area codes which we load now.

#### Read State & Area Code Tables and take a glimpse
```{r LoadCodesBLS}
fac <- system.file("extdata", "BLS_AreaCodes.tsv", package = "dplyrExamples")
area_codes <- read_tsv(fac)
fsc <- system.file("extdata", "BLS_StateCodes.tsv", package = "dplyrExamples")
state_codes <- read_tsv(fsc)
glimpse(area_codes)
glimpse(state_codes)
```

Now we have everything we need to tidy up our data.

#### Tidy Up Area Employment by Month
```{r TidyEmployment}
Employment_By_Area_1995_2015 <- employment %>%
  gather(mmm_yyyy, NonFarm_000, -Series_ID) %>%
  mutate(Month_Of = as.Date(paste0("01_", mmm_yyyy), format = "%d_%b_%Y"),
         state_code = str_sub(Series_ID, 4, 5),
         area_code = str_sub(Series_ID, 6, 10)) %>%
  left_join(state_codes) %>%
  left_join(area_codes) %>%
  mutate_each(funs(factor), ends_with("name")) %>% 
  rename(State = state_name, Area = area_name) %>%
  select(State, Area, Month_Of, NonFarm_000) %>%
  arrange(State, Area, Month_Of)
glimpse(Employment_By_Area_1995_2015)
```

#### Simple EDA analytics with tidy data
Now that we have the employment data is a tidy form, it's easy to use simple analyses:

```{r Simple EDA, fig.width=9, fig.height=10}
summary(Employment_By_Area_1995_2015)

AreasInNM<- Employment_By_Area_1995_2015 %>% 
  filter(State == "New Mexico",
         Month_Of >= as.Date("2004-01-01")) %>% 
  mutate(Year = factor(year(Month_Of)),
         Month = month(Month_Of)) %>% 
  droplevels()
months <- month(seq(as.Date("2000/1/1"), by = "month", length.out = 12), 
                label = TRUE, abbr = TRUE)
ggplot(AreasInNM, aes(Month, NonFarm_000, color = Year)) + 
  geom_point(size = I(1)) + geom_line() + 
  scale_x_continuous(breaks = 1:12, labels = months) + 
  ggtitle("Employment (000) by Year for Areas in New Mexico") + 
  facet_grid(Area ~ ., scales = "free_y" )
```

We hope you have found this example of usint dplyr and tidyr useful. Please send comments and suggestions to Jim at DS4CI.org or leave an issue or pull request at [github](https://github.com/ds4ci/dplyrExamples).

Thanks!
Jim

```{r CleanHouse}
file.remove(file_out)
```

  **END**
