---
title: "dplyr Example 1 - BLS Wide to Tidy"
author: "Jim Porzak"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dplyr Example 1 - BLS Wide to Tidy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The Bureau of Labor Statistics (the BLS) [data page](http://www.bls.gov/data/), has
many data sets and many ways to access them. For a research project we needed the
monthly employment figures by "metro areas." Using the _Multi-screen Data Search_ tool for _Employment, Hours, and Earnings - State and Metro Area_ we captured the non-farm employment from 1995 through 2015 as a tab seperated file. See the BLS _sm_ [file spec](http://download.bls.gov/pub/time.series/sm/sm.txt)

There is a bit clean-up to do to make it readable with `read_tsv()` from Hadley's readr package. See the function `BLS_CleanRawTextFile()`, included in this package, if you are interested in the details - which are not relevant for this dplyr example.

#### Read the BLS file into the data frame `employment` and take a glimpse of the first and last  ten columns

```{r LoadRawBLS}
library(dplyr)
library(tidyr)
library(readr)
library(dplyrExamples)
library(stringr)
library(ggplot2)
library(lubridate)
fn <- system.file("extdata", "BLS_NonFarmEmploymentInAreas_1995_2015.tsv",
                  package = "dplyrExamples")
file_out <- "file_out.tsv"
BLS_CleanRawTextFile(fn, file_out)  ## Clean up the BLS raw text file
employment <- read_tsv(file_out)
dim(employment)
glimpse(employment[1:10])
glimpse(employment[244:253])
```

There are two challenges with this data set:

1. The *Series_ID* needs to be decoded to pull out the state and area for the row. 
1. It is _very_ wide - 253 columns! We wish to tidy it up, in the Hadley sense, so 
each row will just a single numeric column, the number of non-farm employees in the area for the month.

The series_id decoder is in Section 5 of the BLS [SM file spec](http://download.bls.gov/pub/time.series/sm/sm.txt) and is repeated here:

```
0        1         2    <--
12345678901234567890    <-- character counter 
SMU01266207072200001    <-- Sample Series ID

Positions Code		          			Value

1-2       survey abbreviation	=		SM
3         seasonal (code)   	=		U
4-5       state_code	      	=		01
6-10      area_code	        	=		26620
11-12     supersector_code  	=		70
13-18     industry_code	    	=		70722000
19-20     data_type_code  		=		01
```
We did not need NIAC's code breakdown so positions 11-18 are all zeros in our extract.

The BLS has standard lookup files for the state and area codes which we load now.

#### Read State & Area Code Tables and take a glimpse
```{r LoadCodesBLS}
fac <- system.file("extdata", "BLS_AreaCodes.tsv", package = "dplyrExamples")
area_codes <- read_tsv(fac)
fsc <- system.file("extdata", "BLS_StateCodes.tsv", package = "dplyrExamples")
state_codes <- read_tsv(fsc)
glimpse(area_codes)
glimpse(state_codes)
```

Now we have everything we need to tidy up our data.

#### Tidy Up Area Employment by Month

This is the dplyr sequence you would use in production. The next section breaks down the process step-by-step.

```{r TidyEmployment}
Employment_By_Area_1995_2015 <- employment %>%
  gather(mmm_yyyy, NonFarm_000, -Series_ID) %>%
  mutate(Month_Of = as.Date(paste0("01_", mmm_yyyy), format = "%d_%b_%Y"),
         state_code = str_sub(Series_ID, 4, 5),
         area_code = str_sub(Series_ID, 6, 10)) %>%
  left_join(state_codes) %>%
  left_join(area_codes) %>%
  mutate_each(funs(factor), ends_with("name")) %>% 
  rename(State = state_name, Area = area_name) %>%
  select(State, Area, Month_Of, NonFarm_000) %>%
  arrange(State, Area, Month_Of)
glimpse(Employment_By_Area_1995_2015)
```

#### Step-by-step dplyr

##### Convert from wide to narrow using `tidyr::gather()`.
```{r SbS1}
eba <- employment %>% 
  gather(mmm_yyyy, NonFarm_000, -Series_ID)
glimpse(eba)
```
##### Convert character `mmm_yyyy` to Date and pull out state & area codes from Series_ID.
```{r SbS2}
eba <- eba %>% 
mutate(Month_Of = as.Date(paste0("01_", mmm_yyyy), format = "%d_%b_%Y"),
       state_code = str_sub(Series_ID, 4, 5),
       area_code = str_sub(Series_ID, 6, 10))
glimpse(eba)  
```
##### Look-up the state and area names from the codes data frames.
```{r SbS3}
eba <- eba %>% 
  left_join(state_codes) %>%
  left_join(area_codes)
glimpse(eba)
```
##### Convert the names to factors and rename them to user friendly names.
```{r SbS4}
eba <- eba %>% 
  mutate_each(funs(factor), ends_with("name")) %>% 
  rename(State = state_name, Area = area_name)
glimpse(eba)
```
##### Keep just final columns and sort by month within area within state.
```{r SbS5}
eba <- eba %>% 
  select(State, Area, Month_Of, NonFarm_000) %>%
  arrange(State, Area, Month_Of)
glimpse(eba)
```


#### EDA is simple now that we have tidy data!
1. Overall summary.
2. Plot New Mexico monthly employment by year from 2004 through 2015.

```{r Simple EDA, fig.width=9, fig.height=10}
summary(Employment_By_Area_1995_2015)

# pull out New Mexico data
AreasInNM<- Employment_By_Area_1995_2015 %>% 
  filter(State == "New Mexico",
         Month_Of >= as.Date("2004-01-01")) %>% 
  mutate(Year = factor(year(Month_Of)),
         Month = month(Month_Of)) %>% 
  droplevels()

# get month labels for plot
months <- month(seq(as.Date("2000/1/1"), by = "month", length.out = 12), 
                label = TRUE, abbr = TRUE)

ggplot(AreasInNM, aes(Month, NonFarm_000, color = Year)) + 
  geom_point(size = I(1)) + geom_line() + 
  scale_x_continuous(breaks = 1:12, labels = months) + 
  ggtitle("Employment (000) by Year for Areas in New Mexico") + 
  facet_grid(Area ~ ., scales = "free_y" )
```

#### Conclusion

We hope you have found this example of using dplyr and tidyr useful. Please send comments and suggestions to Jim at DS4CI.org or leave an issue or pull request at [my github](https://github.com/ds4ci/dplyrExamples).

Thanks!
Jim

###### P.S. Don't forget to clean house.
```{r CleanHouse}
file.remove(file_out)
```

  **END**
